---
title: "Class 09 Unsupervised Learning Mini-Project"
author: "Cindy Tran"
date: "1/26/2022"
output:
  pdf_document: default
  html_document: default
---

# 1. Exploratory Data Analysis

## Preparing the data

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
head(wisc.df)
```

Let's remove the "diagnosis" column since it gives the answer of whether the cells are malignant or benign. 
```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]

# Create diagnosis vector for later 
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
```

## Exploratory Data Analysis

Explore the data you created before (wisc.data and diagnosis) to answer the following questions:

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```
569 observations in wisc.df

> Q2. How many of the observations have a malignant diagnosis?

```{r}
length(which(diagnosis == 1))
```
212 observations

> Q3. How many variables/features in the data are suffixed with _mean

```{r}
length(grep(pattern = "_mean", x = colnames(wisc.df)))
```
10 variables

# 2. Principal Component Analysis

## Performing PCA

```{r}
# Check column means and standard deviations
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

```{r}
# Perform PCA on wisc.data
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 (PC1, PC2, and PC3)

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 (PC1-7)

## Interpreting PCA Results

```{r}
# Create a biplot of wisc.pr
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is very difficult to interpret because everything is clustered together and hard to distinguish from each other. 

```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[, c(1, 2)], col = (diagnosis + 1),
     xlab = "PC1", ylab = "PC2")
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[, c(1, 3)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC3")
```

There is clearer separation between the 2 subgroups for the first plot (PC1 vs PC2) compared to the second (PC1 vs PC3). These plots are also much neater than the biplot. 

## Variance Explained

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- wisc.df$diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col = diagnosis) + 
  geom_point()
```

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg = paste0("PC",1:length(pve)), las = 2, axes = FALSE)
axis(2, at=pve, labels = round(pve, 2)*100 )
```

```{r}
# ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

## Communicating PCA Results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation[, 1]
```
-0.26085376

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```
5 (PC1-5)

# 3. Hierarchical Clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

data.dist <- dist(data.scaled)

wisc.hclust <- hclust(data.dist, method = "complete")
```

## Results of Hierarchical Clustering

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 20, col="red", lty=2)
```
20

## Selecting Number of Clusters

```{r}
# Cut tree so that it has 4 clusters: wisc.hclust.clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

# Compare cluster membership to actual diagnoses
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
# Cut tree so that it has 3 clusters: wisc.hclust.clusters
wisc.hclust.clusters0 <- cutree(wisc.hclust, k = 3)

# Compare cluster membership to actual diagnoses
table(wisc.hclust.clusters0, diagnosis)
```

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

```{r}
wisc.hclust1 <- hclust(data.dist, method = "single")
plot(wisc.hclust1)

wisc.hclust2 <- hclust(data.dist, method = "average")
plot(wisc.hclust2)

wisc.hclust3 <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust3)
```
I like the ward.D2 method because it separated the results into 2 clear groups. 

# 4. OPTIONAL: K-means Clustering

## K-means Clustering and Comparing Results

```{r}
# Create a k-means model on wisc.data: wisc.km
wisc.km<-kmeans(scale(wisc.data), centers = 2, nstart = 20)

# Compare k-means to actual diagnoses
table(wisc.km$cluster, diagnosis)
```

> Q14. How well does k-means separate the two diagnoses? How does it compare to your hclust results?

Clusters 1, 2, and 4 in the hierarchical clustering model seem to be like cluster 1 from the k-means method. Cluster 3 from the hierarchical clustering model seems to be like cluster 2 from the k-means method. 

```{r}
table(wisc.hclust.clusters, wisc.km$cluster)
```

# 5. Combining Methods

## Clustering on PCA Results

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
plot(wisc.pr.hclust)
```
```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)

table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col = grps)

plot(wisc.pr$x[,1:2], col = (diagnosis + 1))
```

```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
```

```{r}
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col = g)
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

# 6. Sensitivity/Specificity

> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Kmeans resulted in the best specificity and sensitivity. 

# 7. Prediction

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col = g)
points(npc[,1], npc[,2], col = "blue", pch = 16, cex = 3)
text(npc[,1], npc[,2], c(1,2), col = "white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize following up with those patients represented by the black circles overlapping with the red circles because their cells are similar to cells of malignant tumors. 

# About This Document

```{r}
sessionInfo()
```







