---
title: "Class 13 Unix Lab"
author: "Cindy Tran"
date: "3/1/2022"
output: pdf_document
---
# 4. Inspect our Sequences

> Q. What does the star character accomplish here?

It applies the command to anything that ends with what is after the star. 

> Q. How many sequences are in this mouse.1.protein.faa file? Hint: Try using grep to figure this out…

68180

> Q. What happens if you run the above command without the > mm-first.fa part?

It won’t put the output of the command into mm-first.fa

>Q. What happens if you were to use two ‘>’ symbols (i.e. >> mm-first.fa)?

It would add the data to an existing file called mm-first.fa, but since we don’t have a file called that, it would lead to an error.

# 6. Running More BLAST Jobs

> Q.How would you determine how many sequences are in the mm-second.fa file?

Grep -c ‘>’ mm-second.fa

# 10. Using RStudio Online to Read Your Output

```{r}
tsv <- readr::read_delim("mm-second.x.zebrafish.tsv", col_names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore"))
```

Make a histogram of the $bitscore values. You may want to set the optional breaks to be a larger number (e.g. breaks=30).

```{r}
library(ggplot2)
ggplot(tsv) +
  aes(bitscore) +
  geom_histogram()
```
> Q. What do you notice here? Note that larger bitscores are better.

Most bitscores are around 300/ 

> Is there a straightforward relationship between percent identity ($pident) and bitscore ($bitscore) for the alignments we generated?

The answer is that bitscores are only somewhat related to pident; they take into account not only the percent identity but the length of the alignment.

```{r}
plot(tsv$pident  * (tsv$qend - tsv$qstart), tsv$bitscore)
```

```{r}
ggplot(tsv, aes(pident, bitscore)) + geom_point(alpha=0.1) 
```

```{r}
ggplot(tsv, aes((tsv$pident * (tsv$qend - tsv$qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth()
```
## On your local machine
scp -i ~/Downloads/barry_bioinf.pem -r ubuntu@YOUR_IP_ADDRESS:~/work/* .
> Q. Note the addition of the -r option here: What is it’s purpose? Also what about the *, what is it’s purpose here?

-r means to copy the entire directory and * means to copy everything that is in the work directory. 

> Q. Why did it take longer to BLAST mm-second.fa than mm-first.fa?

Because it is a larger file

> Q. When we plot e-values why do you often work in -log(evalue) units?

Because it makes relationships easier to see and smoother lines when there are large variances.  

> What is an advantage of rsync over scp?

Rsync allows you to sync both remote and local directories. 

> What is the advantage of using R (and other tools) on remote machines vs our local computer?

Remote machine can do things much faster. 

> Q. What is the disadvantage of remote vs local work?

It requires more setup to work remotely.
