---
title: "Class 12: Bioinformatics in drug discovery and design (Focus on new AlphaFold2)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.1 Obtaining and inspecting our input structure.

```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
hiv <- read.pdb(file.name)
hiv
```

> Q1: What is the name of the two non protein resid values in this structure? What does resid correspond to and how would you get a listing of all resid values in this structure? 

The 2 non protein residues are HOH and MK1. Resid corresponds to the position of the molecule. You would do attributes(hiv) and then hiv$atom to see the listing of all resid values. 

```{r}
attributes(hiv)
head(hiv$atom)
```

# 1.2 Prepare initial protein and ligand input files

```{r}
prot <- trim.pdb(hiv, "protein")
lig <- trim.pdb(hiv, "ligand")
```

```{r}
______(prot, file="1hsg_protein.pdb")
______(lig, file="1hsg_ligand.pdb")
```

# 1.3 Using AutoDockTools to setup protein docking input









