---
title: "Class 14 Vaccination Rate Mini Project"
author: "Cindy Tran"
date: "2/13/2022"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated?

persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area?

zip_code_tabulation_area

> Q3. What is the earliest date in this dataset?

2021-01-05

> Q4. What is the latest date in this dataset?

2022-02-08

```{r}
#install.packages("skimr")
library(skimr)
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

15

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```
9640

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
9640 / (sum(!is.na(vax$persons_fully_vaccinated))) * 100

```

10%

> Q8. [Optional]: Why might this data be missing?

This data is posisbly missing because some counties did not have collect this information. 

## Working with Dates

```{r}
library(lubridate)
```
```{r}
today()
```

```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now we can do math with dates. For example: How many days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]
```

Using the last and the first date value we can now determine how many days the dataset span.

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP Codes

```{r}
#install.packages("zipcodeR")
library(zipcodeR)
```

Find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code area.
```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles
```{r}
zip_distance('92037','92109')
```

We can pull census data about ZIP code areas (including median household income etc.

```{r}
reverse_zipcode(c('92037', "92109") )
```

```{r}
# Pull data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego Area

```{r}
# Subset to San Diego county only areas
sd <- vax[ vax$county == "San Diego" , ]
nrow(sd)

library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

Using dplyr is often more convenient when we are subsetting across multiple criteria - for example all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
which.max(sd$age12_plus_population)
sd[56,]
```
92154

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?

```{r}
q13 <- filter (sd, as_of_date == "2021-11-09")
mean(q13$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
library(ggplot2)
ggplot(q13) +
  aes(x = percent_of_population_fully_vaccinated) + 
  geom_histogram()
```

## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date, percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

## Comparing 92037 to Other SImilarly Sized Areas

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated, na.rm = TRUE)
```
```{r}
ggplot(ucsd) +
  aes(as_of_date, percent_of_population_fully_vaccinated) + 
  geom_hline(yintercept = 0.6716873, linetype = "dashed", color = "red") +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(x = percent_of_population_fully_vaccinated) +
  geom_histogram()
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
92109 is above the average while 92040 is below the average.

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

mean(vax.36.all$percent_of_population_fully_vaccinated, na.rm = TRUE)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.472364, linetype="dashed")
```

> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I feel a bit uncomfortable going to in person classes, but am okay with going if necessary.  

```{r}
sessionInfo()
```






