---
title: "Class 08 Lab (Hands on with PCA)"
author: "Cindy Tran"
date: "1/25/2022"
output:
  pdf_document: default
  html_document: default
---

# First up, kmeans()

Demo of using a kmeans() function R. First, we'll make up some data with a known structure. 

```{r}
tmp <- c(rnorm(30, -3), rnorm(30, 3))
tmp
x <- cbind(x = tmp, y = rev(tmp))
x
plot(x)
```

Now we have some made-up data in 'x'. Let's see how kmeans() works with this data

```{r}
k <- kmeans(x, centers = 2, nstart = 20)
k
```
> Q. How many points are in each cluster?

```{r}
k$size
```

> Q. How do we get to the cluster membership/assignment?

```{r}
k$cluster
```

> Q. What about cluster centers?

```{r}
k$centers
```

Now we got to the main results. Let's use them to plot our data with the kmeans() result.

```{r}
plot(x, col = k$cluster)
points(k$centers, col = "blue", pch = 15)
```

## Now for Hierachical Clustering

We will cluster the same data 'x' with the 'hclust()'. In this case, 'hclust()' requires a distance matrix as input. 

```{r}
hc <- hclust(dist(x))
hc
```

Let's plot our hclust() result

```{r}
plot(hc)
```

To get our cluster membership vector, we need to "cut" the dendrogram tree with 'cutree()' 

```{r}
grps <- cutree(hc, h = 8)
grps
```

Now plot our data with the hclust() results. 

```{r}
plot(x, col = grps)
```


# Principal Component Analysis (PCA)

## PCA of UK Food Data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named 'x'? What R functions could you use to answer this question?

```{r}
nrow(x)
ncol(x)
```
17 rows and 5 columns. 

Preview the first 6 rows

```{r}
head(x)
```
```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
Checking the dimensions again

```{r}
dim(x)
```
17 rows, 4 columns

> Q2. Which approach to solving the 'row-names problem' above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the 'x <- read.csv(url, row.names = 1)' because if you run code block 'x <- x[,-1]' multiple times, it will change what column will be your column 1.

```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x), col = cols)
```

```{r}
barplot(as.matrix(x), col = cols, beside = TRUE)
```

> O3. Changing what optional argument in the barplot() function results in stacked bars? 

beside = FALSE

> Q5. Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col = cols)
```

The closer the points are on the diagonal, the more similar the variables are to each other. If a given point lies on the diagonal, that means that the 2 variables being compared have the same values. 

> Q6 What is the main difference between N. Ireland and the other countries of the UK in terms of this dataset?

Northern Ireland is more different from all of the other UK countries than those countries are from each other. 

PCA to the rescue!!

The main base PCA function is called 'prcomp()'. We will need to give it the transpose of our input data. 

```{r}
pca <- prcomp(t(x))
```

There is a nice summary of how well PCA is doing. 

```{r}
summary(pca)
```

```{r}
attributes(pca)
```

To make our new PCA plot (a.k.a. PCA score plot), we access 'pca$x'

> Q7. Generate a plot of PC1 vs PC2.

```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at the start of this document. 

Color up the plot

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2", xlim = c(-270, 500))
text(pca$x[,1], pca$x[,2], colnames(x), col = country_cols)
```

Calculating how much variation in the original data each PC accounts for. Take the square of pcs$dev, which stands for "standard deviation."

```{r}
v <- round (pca$sdev^2/sum(pca$sdev^2) * 100)
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

Digging Deeper (variable loadings)

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

> Q9. Generate a similar 'loadings plot' for PC2. What two food groups feature prominently, and what does PC2 mainly tell us about?

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

Fresh_fruit and Alcoholic_drinks feature most prominently. PC2 accounts for the second-most amount of variation of the data. 

Biplots

```{r}
biplot(pca)
```


## PCA of RNA-seq Data

Read in data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

> Q10. How many genes and samples are in this data set?

```{r}
nrow(rna.data)
```

```{r}
pca <- prcomp(t(rna.data), scale = TRUE)
summary(pca)
```

Do our PCA plot of this RNA-Seq data

```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
```

```{r}
plot(pca, main = "Quick scree plot")
```

Making the above scree plot ourselves

```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```
```{r}
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```


```{r}
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
text(pca$x[,1], pca$x[,2], colnames(rna.data))
```

```{r}
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

Using ggplot

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```

Optional: Gene Loadings

Finding the top 10 measurements (genes) that contribute most to pc1 in either direction (+ or -)

```{r}
loading_scores <- pca$rotation[,1]

# Find the top 10 measurements (genes) that contribute most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

# show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```




















